Trade Dependence and Militarized Interstate Disputes

### 2022-12-13 article/article.qmd
Update output to pdf

---
title: "Trade Dependence"
format: 
  html:
    self-contained: true
bibliography: bibliography.bib
execute: 
  message: false
  warning: false
  echo: false
---

@@

  theme_minimal() + 
  theme(plot.title.position = "plot",
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold"),
        legend.position = "bottom")

@@

Does trade dependence reduce conflict between states? Debate remains over the role and significance of economic interdependence in promoting peace: different scholars have found that dependence increases, decreases, and has no significant effect on the likelihood of conflict between states. However, this research has generally relied on a broad measure of bilateral trade dependence. This paper will delve deeper into the effect of this economic relationship by exploring product-level dependence. If trade dependence deters conflict between states by increasing the expected cost of conflict through trade disruptions, we should measure the true cost of this disruption. Rarely, if ever, do we see states threaten to or impose total trade embargoes. Rather, states impose trade bans on specific products that maximize the cost borne by their opponent while minimizing the cost to their own economy. Using a more disaggregated measure of trade dependence, I find that dependence does not have a significant effect on the likelihood of militarized disputes between states.

@@

Debate remains over the role played by, and significance of, bilateral trade dependence in conflict mitigation. Recent scholarship on the topic has begun exploring whether this ambiguity is the product of measurement error. Within international relations scholarship, measures of trade dependence have been defined at a broad level, predominantly using a measure introduced by Oneal and Russett [-@oneal1997] that explores total trade between two states relative to their gross domestic product (GDP). However, a more disaggregated approach to trade dependence is worth exploring. We are interested in understanding how trade dependence can deter conflict. Rarely (if ever) do states threaten to sever their entire trade relationship with another state, including trade dependent states. Rather, we regularly see states threaten or enact trade bans on particular products. In fact, states often aim to threaten to, or actually, inflict the greatest harm on their rival whilst minimizing the costs they themselves will face. A broad understanding of trade dependence does not capture the real threat a potential aggressor faces or should expect to face. A more accurate measure of this factor is product-level trade dependence. Asymmetries at this level have been used as a foreign policy mechanism to deter or punish rival states.

Before we explore this dynamic in detail, let us step out in detail the process we seek better understand. First, tensions between two states build. This could be for a number of reasons, including historical animosity, a diplomatic spat over a regional or global issue, or domestic political interests prompting an unfavourable bilateral foreign policy. The threatened state issues a formal request to the initiating state to change or reverse their policy actions. The initiating state refuses: the incentives for enacting the controversial policy continue to outweigh the cost of a soured relationship with the threatened state. The threatened state then states that it will adopt punitive policies towards the initiator, including trade embargoes on specific products on which the initiator depends. Here, the initiator's expected costs from pursuing this policy change: they know that if they continue along this path they will likely face economic costs in the form of trade disruptions. Does this deter them? Many scholars would ask: how costly is this disruption? This paper will answer this question. If the costs are not so great, the initiator may respond to this threat through further escalation, including through a militarized dispute. International relations scholarship uses this broad measure of conflict to include actions that sit below invasion or war. This is often the precursor to full-scale interstate war. States use these actions as an extreme signal of their intent to escalate a dispute to war, hoping to deter their opponent.

@@

This analysis focuses on all directed state dyads in 2014. This includes `r nrow(full_df) |> comma()` observations. We are interested in whether or not a trade dependent state initiates conflict against another state. This is a directed question. We need to know whether the potential initiator is dependent on their opponent, not simply whether either partner has a dependence on their other.

@@

Following other literature to enhance comparability, I include the following controls.

@@

We will compare this to the predicted probability derived using the traditional broad measure of trade dependence.

```{r}
tbl_merge(list(t1, t2), tab_spanner = c("Product level", "Broad level"))
```

@@

*Figure 1* illustrates the different predicted probabilities that state $i$ will initiate an MID against state $j$ for each possible level trade dependence between no dependence (0 percent) and total dependence (100 percent).[^3]

[^3]: Any prediction made for a level of trade dependence above `r max(full_df$trade_dep_total) |> percent()` for product-level dependence and `r max(full_df$trade_dep_broad) |> percent()` for total trade dependence is an out of sample prediction.

```{r}
ggplot(pred_m1_m2, aes(x = trade_dep, y = .fitted, colour = type)) + 
  geom_line() + 
  theme_custom() + 
  labs(
    title = "Predicted probability of MID given product-level trade dependence",
    x = "Trade dependence (%)",
    y = NULL,
    colour = "Measure"
  )
```

@@

Moving from these base models to ones that incorporate traditional controls for MIDs, we learn that the relationship between conflict and trade dependence is not significant.

```{r}
tbl_merge(list(t3, t4), tab_spanner = c("Product level", "Broad level"))
```

For both product-level and total trade dependence, the odds that state $i$ will initiate an MID against state $j$ remain positive; however, the statistical significance of this relationship drops well below traditional thresholds. Significant determinants of conflict include contiguity and state $i$'s level of democracy. Concentrating on product-level trade dependence, I find that state $i$ is less to initiate an MID against a state with which it shares a land border. I also find that as state $i$ becomes more democratic, it is less likely to initiate a MID against state $j$. Interestingly, I find that state $j$'s level of democracy does not appear to be a significant factor in state $i$'s decision to initiate an MID, holding all else constant.

>>>>

---
title: "Trade Dependence and Militarized Interstate Disputes"
format: 
  pdf:
    documentclass: article
    papersize: letter
    geometry:
      - margin=1in
    linestretch: 2
bibliography: bibliography.bib
execute: 
  message: false
  warning: false
  echo: false
---

@@

  theme_minimal() + 
  theme(plot.title.position = "plot",
        panel.grid = element_blank(),
        plot.title = element_text(face = "bold"),
        legend.position = "bottom",
        text = element_text(family = "serif"))

@@

Does trade dependence reduce conflict between states? Debate remains over the role and significance of economic interdependence in promoting peace: different scholars have found that dependence increases, decreases, and has no significant effect on the likelihood of conflict between states. However, this research has generally relied on a broad measure of bilateral trade dependence. This paper will delve deeper into the effect of this economic relationship by exploring product-level dependence. If trade dependence deters conflict between states by increasing the expected cost of conflict through trade disruptions, we should measure the true cost of this disruption. Rarely, if ever, do we see states threaten to or impose total trade embargoes. Rather, states impose trade bans on specific products that maximize the cost borne by their opponent while minimizing the cost to their own economy. Using a more disaggregated measure of trade dependence, I find that this dependence does not have a significant effect on the likelihood of militarized disputes between states.

@@

Debate remains over the role played by, and significance of, bilateral trade dependence in conflict mitigation. Recent scholarship on the topic has begun exploring whether this ambiguity is the product of measurement error. Within international relations scholarship, measures of trade dependence have been defined at a broad level, predominantly using a measure introduced by Oneal and Russett [-@oneal1997] that explores total trade between two states relative to their gross domestic product (GDP). However, a more disaggregated approach to trade dependence is worth exploring. We are interested in understanding how trade dependence can deter conflict. Rarely (if ever) do states threaten to sever their entire trade relationship with another state. Rather, we regularly see states threaten or enact trade bans on particular products. In fact, states often threaten to inflict the greatest harm on their rival whilst minimizing the costs they will face. A broad understanding of trade dependence does not capture the real threat a potential aggressor faces. A more accurate measure of this factor is product-level trade dependence. Asymmetries at this level have been used as a foreign policy mechanism to deter or punish rival states.

Before we explore this dynamic in detail, let us step out in detail the process we seek better understand. First, tensions between two states builds. This could be for a number of reasons, including historical animosity, a diplomatic spat over a regional or global issue, or domestic political interests prompting an controversial bilateral foreign policy. The threatened state issues a formal request to the initiating state to change or reverse their policy actions. The initiating state refuses: the incentives for enacting the controversial policy continue to outweigh the cost of a soured relationship with the threatened state. The threatened state responds by threatening to adopt punitive policies towards the initiator, including trade embargoes on specific products on which the initiator depends. Here, the initiator's expected costs from pursuing this policy change: they know that if they continue along this path they will likely face economic costs in the form of trade disruptions. Does this deter them? Many scholars would ask: how costly is this disruption? This paper will answer this question. If the costs are not so great, the initiator may respond to this threat through further escalation, including through a militarized dispute. International relations scholarship uses this broad measure of conflict to include actions that sit below invasion or war. This is often the precursor to full-scale interstate war. States use these actions as an extreme signal of their intent to escalate a dispute to war, hoping to deter their opponent.

@@

This analysis focuses on all directed state dyads in 2014. This includes `r nrow(full_df) |> comma()` observations. We are interested in whether or not a trade dependent state initiates conflict against another state. This is a directed question. We need to know whether the potential initiator is dependent on their opponent, not simply whether either partner has a dependence on the other.

@@

Following other literature exploring this question, I include several control variables. I include a binary variable accounting for whether the dyad includes states that share a land border. I also include two variables providing each state's level of democracy.

@@

First, let us compare the relationship between MID initiation and product-level trade dependence compared to the relationship estimated using the more traditional, broad measure of trade dependence.

```{r}
#| tbl-cap: Base models

tbl_merge(list(t1, t2), tab_spanner = c("Product level", "Broad level"))
```

@@

*Figure 1* illustrates the different predicted probabilities that state $i$ will initiate an MID against state $j$ for each possible level trade dependence between no dependence (0 percent) and total dependence (100 percent).[^2]

[^2]: Any prediction made for a level of trade dependence above `r max(full_df$trade_dep_total) |> percent()` for product-level dependence and `r max(full_df$trade_dep_broad) |> percent()` for total trade dependence is an out of sample prediction.

```{r}
#| fig-cap: Predicted probability of MID given product-level trade dependence
ggplot(pred_m1_m2, aes(x = trade_dep, y = .fitted, colour = type)) + 
  geom_line() + 
  theme_custom() + 
  labs(
    x = "Trade dependence (%)",
    y = NULL,
    colour = "Measure"
  )
```

@@

Moving from these base models to ones that incorporate traditional controls for MIDs and trade relationships, we learn that the relationship between conflict and trade dependence is not significant.

```{r}
#| tbl-cap: Controlled models

tbl_merge(list(t3, t4), tab_spanner = c("Product level", "Broad level"))
```

For both product-level and total trade dependence, the odds that state $i$ will initiate an MID against state $j$ remain positive; however, the statistical significance of this relationship drops well below traditional thresholds. Significant determinants of conflict include contiguity and state $i$'s level of democracy. I find that state $i$ is less to initiate an MID against a state with which it shares a land border, on average holding all else constant. I also find that as state $i$ becomes more democratic, it is less likely to initiate a MID against state $j$. Interestingly, I find that state $j$'s level of democracy does not appear to be a significant factor in state $i$'s decision to initiate an MID.

## Bibliography
